### 2018-06-29 contents/reviews/pain.tex
final submission

The gate-control theory \cite{Melzack1965a} (Figure \ref{fig:gate-theory}), which forms the basis of modern theory of pain, postulates that the cells of the substantia gelatinosa selectively inhibit signal transmission from primary afferent to the central transmission cells in the spinal dorsal horn that relays touch or pain sensations. The transmission control depends on the large and small diameter primary afferents, where large fibres inhibit, while small fibres facilitate the transmission. The gate-control theory reconciles the two previously divergent specificity and pattern theories of pain.

@@

An important contribution of the gate-control theory was to emphasize central nervous system mechanisms \cite{Melzack1999}, and the role of the dorsal horn in pain processing, and therefore reconciles the evidences of pain-specific primary afferent and with pattern theory. Contemporary views consider pain to be complex and multi-dimensional, where the processing of ascending nociceptive signals, in conjunction with descending modulatory pathways, shape the pain experience. Melzack and Casey further proposed that pain consists of sensory-discriminative, affective-motivational, and cognitive-evaluative components \cite{Melzack1968}. Specifically, the sensory-discriminative component consists of peripheral nociceptive afferents transmit to the spinal dorsal horn (laminae V), whereby nociceptive specific (NS) and wide-dynamic-range (WDR) neurons convergently process these primary afferents, and relay them to the thalamus, and subsequently the primary sensory cortex. A central concept, which was also introduced by Melzack, is the  neuromatrix theory of pain, which dictates that the pain experience is produced by a "neurosignature" pattern, that is generated from widely distributed neural network, which Melzack refers as the "body-self neuromatrix". \cite{Melzack1999}. The concept is now more commonly referred to as the "pain matrix" \cite{Tracey2007a}. (See section \ref{section:neuromatrix})

@@

\textit{Nociceptors}, the peripheral nociceptors respond to noxious stimuli. These receptors have free nerve endings, and can respond to thermal, mechanical, and/or chemical stimuli, such as physical trauma or heat/cold temperatures. These receptors broadly separate into two categories, differentiated by the fibre type. Acute nociceptors are associated with $A\delta$ fibres, which permit fast signal conductance (5 to 30 m/s) and trigger a prickly or sharp pain sensation; examples include needle prick other acute trauma. Dull pain is triggered by receptors that are associated with slow conducting C fibres ($<1$ m/s). C-fibre nociceptors can be further differentiated into mechano-heat, silent, and polymodal subtypes. $A\delta$-fibre nociceptors can be divided into high-threshold mechano-heat, low-threshold mechano-heat, and noxious cold subtypes \cite{Julius2013}.

@@

\begin{figure}[ht]
\includegraphics[width=0.8\textwidth]{dorsal-lamina.pdf}
\centering
\caption{The axonal projections of the fast conduction ($ A\delta$) and slow conducting (C) pain stimuli in the dorsal horn. Modified from \protect\cite{DMello2008}}
\label{fig:drg}
\end{figure}

The interneurons in the dorsal horn plays an important part in somatosensory integration as the sensory afferent pass from peripheral into the central nervous system. For example, the wide dynamic range (WDR) neurons are found primarily in the spinal dorsal horn in lamina V-VI. These neurons respond to both innocuous and noxious stimuli, and are decidedly non-specific. They are found to sustain their impulse discharge during long durations of repetitive nociceptive stimulation, while nociceptive specific (NS) neurons showed a declined discharge rate within the same duration \cite{Coghill1993,Maixner1986}, thereby providing evidence that WDR neurons are responsive to prolonged nociceptive pain.

In each spinal afferent level, the dorsal horn neurons project noceciptive signals to the brain via five distinct paths (Figure \ref{fig:pain-pathways}). Lamina I, V-VII axons crosses the mid-line to form the \textit{spinothalamic} tract that ascend in the contralateral anterolateral column of the spinal cord, and project to thalamic nuclei. Laminae VII and VIII neurons from the \textit{spinorecticular} tract ascend via the anterolateral quadrant of the spinal cord, without crossing the midline, to reach the reticular formation and thalamus. The \textit{spinomesencephalic} tract include axons from laminae I and V, and ascend in the anterolateral quadrant to the mesencephalic reticular formation, including the periaqueductual gray. The \textit{cervicothalamic} tract include only the white matters of the upper two cervical segments. It takes inputs from laminae III and IV, and projects to the thalamus via the medial lemniscus. The spinohypothalamic tract contains axons from laminae I, V, and VIII, and project to hypothalamic nuclei for autonomic functions \cite{kandel2000principles}.

@@

Before reaching the neocortex, several important second order pain afferents arrive at the thalamus. The spinothalamic tracts subserving face and body sensory information terminate in several regions of the thalamus. A major termination site is the ventroposterior complex (VP, also known as ventrocaudal (Vc) nucleus), where ventroposterior-medial (VPM) nuclei and ventroposterior-lateral (VPL) nuclei are termination sites for the face and body respectively. The VP complex sends third-order afferents to the primary sensory cortex (S1). Thus, the VP representation is somatotopically organized with the face region in the medial aspect, and lower limbs towards the lateral aspect. The representation is retained in the cortical projection, where the S1 somatotopy is shown to have the face region towards the lateral surface of the S1 cortical strip, and lower limb towards the superior-medial aspect.

@@

Pain percept, defined as the subjective experience of pain, is thought to involve distributed regions of the brain, coined as the pain “neuromatrix”\cite{Melzack1999a}, or more commonly the "pain matrix". The regions are not functionally independent, but rather thought to be dynamically inter-related.

 Important cortical regions involved in pain include anterior and middle cingulate cortices (ACC/MCC), insula, prefrontal cortex (PFC), thalamus, periaqueductal grey (PAG), secondary somatosensory cortex (SII) and motor cortex (M1). The role of these cortical regions in pain were studied in neuroimaging fMRI studies \cite{Apkarian2013c,Wager2013,Davis2012a}, which demonstrated consistent BOLD co-activations during noxious stimulation. (Figure \ref{fig:pain-matrix}).

@@

A recent study suggests that pain is correlated to the dynamically changing patterns of BOLD activities over time in the regions involved. Thus, the focus of the salient pain percept requires a switch away from the default mode network \cite{Kucyi2013}. The dynamic rhythms of activities that attune to and away from saliency, form a dynamic pain connectome \cite{Kucyi2015}. Chronic pain therefore demonstrated differential tuning of the default mode network and attention patterns of the brain due to their pathological activation of the pain matrix in the absence of detectable noxious stimuli \cite{Baliki2008,Legrain2009}.

\paragraph{Sensory-Discrimination}
Cortical regions associated with sensory-discriminative aspect of pain, such as the location and intensity of pain, include the S1 and S2 cortices. The S1 cortex is responsible for the discrimination of fine-touch sensations. S1 receives inputs directly from VP thalamus. Nonetheless, S1 has not been consistently reported in acute pain studies, possibly due to the difficulty in detecting pain-related activations as there are relatively small number of nociceptive neurons within S1 \cite{bushnell1999pain}. S2, however, receives direct nociceptive input from VPI thalamus, and its BOLD activation is more consistently observed across pain fMRI studies \cite{Wager2013,Brodersen2012}. The posterior insula also receives direct nociceptive inputs from the VMpo thalamus \cite{Craig2003}, and shows strong functional and structural connectivity with S2 \cite{Wiech2014}. Therefore the insula, as well as S1, are believed to plays an important role in sensory-discrimination aspect of pain.

@@

\paragraph{Default mode network}
The default mode network (DMN) are regions of the brain, including the cingulate cortex, which show highly correlative BOLD activations in fMRI during wakeful rest \cite{Greicius2003,Raichle2015}. The regions have been shown to consistently decrease in activity during attention-demanding, and non-self-referential tasks,when compared with activities during relaxed non-task states \cite{Raichle2015}. The DMN is believed to instantiate processes that support self-referential activities, emotional processing, and recollection \cite{Raichle2015}. The core regions associated with the default network are ventral medial prefrontal cortex (vmPFC), dorsal medial prefrontal cortex (dMPFC), posterior cingulate, inferior parietal, lateral temporal cortex, and hippocampal formation. Regions such as vmPFC and dMPFC overlap with the affective-motivational pain matrix, and are found to be disrupted in DMN in chronic pain \cite{Baliki2008}. Pain related DMN deactivation have been found to be attenuated  in mind wandering away from pain, and may reflect antinociceptive processes \cite{Kucyi2013}, and that the DMN may undergo functional reorganization under chronic pain \cite{Baliki2014}.

@@

The insula is an important neuroanatomical structure that has been implicated in both acute \cite{Lorenz2005,Starr2009} and chronic pain \cite{May2008,Jensen2016}. Anatomically it is subdivided into anterior and posterior components by the central insular sulcus. It can be further identified by its gyral features, consisting of 5 major gyri: anterior, middle and posterior short gyri in the anterior insula, and anterior long, posterior long gyri of the posterior insula. These gyri are not found consistently in the general population, but rather show individual variability \cite{Rosen2015}. Craig postulates that the insula is a key region for homeostasis and interoception \cite{Craig2003b}, and is necessary for pain discriminative sensations via VMpo projections in the middle insula in humans \cite{Craig2003a}. The anterior insula has been shown by functional connectivity analysis to have interactions with anterior MCC and middle insula with posterior MCC \cite{Taylor2009}. Further suggest that the anterior insula and MCC are involved in an emotional/salience network for interoception and emotional integration and monitoring \cite{Vogt2016}.

@@

Neuropathic pain can also arise from changes in the central neuroaxis. Trauma and other disruptions to the spinal pathways can affect noxious signal transmission. Where repeated and sustained noxious stimuli may cause the central nervous system to adapt and amplify responses to stimulation. These can involve activation-dependence plasticity in neurons of the dorsal horn, which result in windup of action potential discharge of these neurons. Additionally, long-term depression may also result in the inhibitory mechanism in interneurons of the substantia gelatinosa \cite{Woolf2000}. White matter pathways within the brain may also become damaged by conditions such as trauma, infections, and multiple sclerosis. It is believed that injury to white matter fibres can affect action potential conduction between cortical and subcortical structures. This may give rise or contribute to neuropathic pain \cite{Geha2008}.

@@

 Neuropathic and acute nociceptive pain have a high degree of overlap in brain activations \cite{Moisset2007}, therefore differentiating between neuropathic and acute pain from neuroimaging continue to be an active area of research. Brain regions that are involved in the somatosensation of pain process are involved in both acute and chronic pain. These regions include the posterior insula, lateral thalamus and dorsal striatum. In chronic pain conditions however, there is increased activation in the anterior insula, ventral striatum, orbital frontal cortex and dorsal lateral prefrontal cortex \cite{Jensen2016,Alomar2016,Lorenz2005}. In chronic back pain, diffusion changes in fractional anisotropy was correlated with medial prefrontal cortex and nucleus accumbens connectivity in pain recovery, but not when the pain persistented \cite{Mansour2013}. Evidence also suggest that back pain chronification seem to increase activation in the emotional regions (amygada and medial prefrontal cortex) of the brain, and transition away from sensory region activations \cite{Hashmi2013}.
 Additionally, evidence of a "pain memory" circuitry involving the MCC \cite{Vogt2016,Nevian2017}, and abnormalities in the reward circuitry \cite{Tracey2017} have been implicated in chronic pain.

>>>>

The gate-control theory \cite{Melzack1965a} (Figure \ref{fig:gate-theory}), which contributes to the modern theory of pain, postulates that cells of the substantia gelatinosa selectively inhibits signal transmission from primary afferent to the central transmission cells in the spinal dorsal horn that integrate and relay somatosensation and pain. The transmission control depends on the large ($A\delta$ fibres)  and small diameter (C-fibres) primary afferents, where large fibres inhibit, while small fibres excite the the transmission cell. The gate-control theory reconciles the two previously divergent specificity and pattern theories of pain.

@@

An important contribution of the gate-control theory was to emphasize central nervous system mechanisms \cite{Melzack1999}, and the role of the dorsal horn in pain processing, and therefore reconciles the evidences of pain-specific primary afferents with pattern theory. Melzack and Casey further proposed that pain consists of sensory-discriminative, affective-motivational, and cognitive-evaluative components \cite{Melzack1968}. The sensory-discriminative component includes peripheral nociceptive afferents transmitting to the spinal dorsal horn (laminae V), whereby nociceptive specific (NS) and wide-dynamic-range (WDR) neurons convergently process these primary afferents, and relay them to the thalamus, and eventually the primary sensory cortex. A central concept, also introduced by Melzack, is the  neuromatrix theory of pain, which dictates that the pain experience is produced by a "neurosignature" pattern, that is generated from widely distributed neural network, which Melzack refers as the "body-self neuromatrix". \cite{Melzack1999}. The concept is now more commonly referred to as the "pain matrix" \cite{Tracey2007a}, although the term itself is controversial \cite{iannetti2010neuromatrix} (See section \ref{section:neuromatrix}). Contemporary views consider pain to be complex and multi-dimensional, where the processing of ascending nociceptive signals, in conjunction with descending modulatory pathways, the default mode and salience network, along with temporal changes in these networks \cite{Kucyi2015}, shape the pain experience.

@@

Nociceptors are peripheral receptors that respond to noxious stimuli. These receptors have free nerve endings, and can respond to thermal, mechanical, and/or chemical stimuli, such as physical trauma or heat/cold temperatures. These broadly separate into two categories, differentiated by the fibre type. Nociceptors responsible for acute pain are associated with $A\delta$ fibres, which permit fast signal conductance (5 to 30 m/s) and trigger a prickly or sharp pain sensation; examples include needle prick other acute trauma. Dull pain is triggered by receptors that are associated with slow conducting C fibres ($<1$ m/s). \cite{Julius2013}.

@@

\begin{figure}[ht]
\includegraphics[width=0.8\textwidth]{dorsal-lamina.pdf}
\centering
\caption{The axonal projections of the fast conducting ($ A\delta$) and slow conducting (C) pain stimuli in the dorsal horn. Modified from \protect\cite{DMello2008}}
\label{fig:drg}
\end{figure}

The interneurons in the dorsal horn plays an important part in somatosensory integration as the sensory afferent pass from peripheral into the central nervous system. For example, the wide dynamic range (WDR) neurons are found primarily in the spinal dorsal horn in lamina V-VI. These neurons appear to respond to both innocuous and noxious stimuli, and are non-specific. They are found to sustain their impulse discharge during long durations of repetitive nociceptive stimulation, while nociceptive specific (NS) neurons showed a declined discharge rate within the same duration \cite{Coghill1993,Maixner1986}, thereby providing evidence that WDR neurons are responsive to prolonged nociceptive pain.

In each spinal afferent level, the dorsal horn neurons project noceciptive signals to the brain via five distinct paths (Figure \ref{fig:pain-pathways}). Lamina I, V-VII axons crosses the mid-line to form the \textit{spinothalamic} tract that ascend in the contralateral anterolateral column of the spinal cord, and project to thalamic nuclei. Laminae VII and VIII neurons from the \textit{spinorecticular} tract ascend via the anterolateral quadrant of the spinal cord, without crossing the midline, to reach the reticular formation and thalamus. The \textit{spinomesencephalic} tract include axons from laminae I and V, and ascend in the anterolateral quadrant to the mesencephalic reticular formation, including the periaqueductual gray. The spinohypothalamic tract contains axons from laminae I, V, and VIII, and project to hypothalamic nuclei for autonomic functions \cite{kandel2000principles}.

@@

Before reaching the neocortex, several important second order pain afferents arrive at the thalamus. The spinothalamic tracts subserving face and body sensory information terminate in several regions of the thalamus. A major termination site is the ventroposterior complex (VP, also known as ventrocaudal (Vc) nucleus), where ventroposterior-medial (VPM) nuclei and ventroposterior-lateral (VPL) nuclei are termination sites for the face and body respectively. The VP complex sends third-order afferents to the primary sensory cortex (S1). Thus, the VP representation is somatotopically organized with the face region in the medial aspect, and lower limbs towards the lateral aspect. The representation is retained in the cortical projection, where the S1 somatotopy is shown to have the face region towards the lateral surface of the S1 cortical strip, and lower limb towards the superior-medial aspect \cite{kandel2000principles}.

@@

Pain percept, which is defined as the subjective experience of pain, is thought to involve distributed regions of the brain that are commonly referred to as the pain “neuromatrix”\cite{Melzack1999a}, or more commonly the "pain matrix". The regions are not functionally independent, but rather thought to be dynamically inter-related \cite{Kucyi2015}. As such, multiple neuroanatomical areas as well as networks are implied in the experience of pain.

 Important cortical regions involved in pain, which demonstrated consistent BOLD co-activations during noxious stimulation, include anterior and middle cingulate cortices (ACC/MCC), insula, prefrontal cortex (PFC), thalamus, periaqueductal grey (PAG), secondary somatosensory cortex (SII) and motor cortex (M1). The role of these cortical regions in pain have been further investigated using neuroimaging fMRI studies \cite{Apkarian2013c,Wager2013,Davis2012a}. (Figure \ref{fig:pain-matrix}).

@@

A recent study suggests that pain is correlated to the dynamically changing patterns of BOLD activities over time in the regions involved. Thus, the focus of the pain percept requires a switch away from the default mode network \cite{Kucyi2013}. The dynamic rhythms of activities that attune to and away from pain, form a dynamic pain connectome \cite{Kucyi2015}. Chronic pain demonstrated differential functional activation of the default mode network and attention patterns of the brain \cite{Baliki2008,Legrain2009}.

\paragraph{Sensory-Discrimination}
Cortical regions associated with sensory-discriminative aspect of pain, such as the location and intensity of pain, include the S1 and S2 cortices. The S1 cortex is responsible for the discrimination of fine-touch sensations. S1 receives inputs directly from VP thalamus. Nonetheless, S1 has not been consistently reported in acute pain studies, possibly due to the difficulty in detecting pain-related activations as there are relatively small number of nociceptive neurons within S1 \cite{bushnell1999pain}. S2, however, receives direct nociceptive input from VPI thalamus, and its BOLD activation is more consistently observed across pain fMRI studies \cite{Wager2013,Brodersen2012}. The posterior insula also receives direct nociceptive inputs from the VMpo thalamus \cite{Craig2003}, and shows strong functional and structural connectivity with S2 \cite{Wiech2014}. Therefore the posterior insula, as well as S1, are believed to plays an important role in sensory-discrimination aspect of pain.

@@

\paragraph{Default mode network}
The default mode network (DMN) are regions of the brain which show highly correlative BOLD activations in fMRI during wakeful rest \cite{Greicius2003,Raichle2015}. The core regions associated with the default network are ventral medial prefrontal cortex (vmPFC), dorsal medial prefrontal cortex (dMPFC), posterior cingulate, inferior parietal, lateral temporal cortex, and hippocampal formation. The regions have been shown to consistently decrease in activity during attention-demanding, and non-self-referential tasks,when compared with activities during relaxed non-task states \cite{Raichle2015}. The DMN is believed to instantiate processes that support self-referential activities, emotional processing, and recollection \cite{Raichle2015}. Regions such as vmPFC and dMPFC overlap with the affective-motivational pain matrix, and are found to be disrupted in DMN in chronic pain \cite{Baliki2008}. Pain related DMN deactivation have been found to be attenuated  in mind wandering away from pain, and may reflect antinociceptive processes \cite{Kucyi2013}, and that the DMN may undergo functional reorganization under chronic pain \cite{Baliki2014}.

@@

The insula is an important neuroanatomical structure that has been implicated in both acute \cite{Lorenz2005,Starr2009} and chronic pain \cite{May2008,Jensen2016}. Anatomically it is subdivided into anterior and posterior components by the central insular sulcus. It can be further identified by its gyral features, consisting of 5 major gyri: anterior, middle and posterior short gyri in the anterior insula, and anterior long, posterior long gyri of the posterior insula. These gyri are not found consistently in the general population, but rather show individual variability \cite{Rosen2015}. Craig postulates that the anterior insula is a key region for homeostasis and interoception \cite{Craig2003b}, and is necessary for pain discriminative sensations via VMpo projections in the middle insula in humans \cite{Craig2003a}. The anterior insula has been shown by functional connectivity analysis to have interactions with anterior MCC and middle insula with posterior MCC \cite{Taylor2009}. Further suggest that the anterior insula and MCC are involved in an emotional/salience network for interoception and emotional integration and monitoring, and posterior insula are involved in pain discrimination \cite{Vogt2016}.

@@

Neuropathic pain can also arise from changes in the central nervous system. Trauma and other disruptions to the periphery or the spinal pathways can affect noxious signal transmission, where repeated and sustained noxious stimuli may cause the central nervous system to adapt and amplify responses to stimulation. These can involve activation-dependence plasticity in neurons of the dorsal horn, which result in windup of action potential discharge of these neurons. Additionally, long-term depression may also result in the inhibitory mechanism in interneurons of the substantia gelatinosa \cite{Woolf2000}. White matter pathways within the brain may also become damaged by conditions such as trauma, infections, and multiple sclerosis. This may give rise or contribute to neuropathic pain \cite{Geha2008}.

@@

 Neuropathic and acute nociceptive pain have a high degree of overlap in brain activations \cite{Moisset2007}, therefore differentiating between neuropathic and acute pain from neuroimaging continue to be an active area of research. In chronic pain conditions, there is increased activation in the anterior insula, ventral striatum, orbital frontal cortex and dorsal lateral prefrontal cortex \cite{Jensen2016,Alomar2016}. In chronic back pain, diffusion changes in fractional anisotropy was correlated with medial prefrontal cortex and nucleus accumbens connectivity in pain recovery, but not when the pain persistented \cite{Mansour2013}. Evidence also suggest that back pain chronification is associated with increase activation in the emotional regions (amygada and medial prefrontal cortex) of the brain, and transition away from sensory region activations \cite{Hashmi2013}.
 Additionally, evidence of a "pain memory" circuitry involving the MCC \cite{Vogt2016,Nevian2017}, and abnormalities in the reward circuitry \cite{Tracey2017} have been implicated in chronic pain.

### 2018-06-29 contents/reviews/trigeminal.tex
Updates from ShareLaTeX
final submission

Modern pain literature continues to update the exact diagnostic definition of TN. The Burchiel classification scheme divides facial pain into 5 major subtypes \cite{Burchiel2003,Eller2005}: idiopathic (without apparent cause), trigeminal injury, symptomatic, postherpetic, and atypical facial pain. Idiopathic TN has 2 sub-types: 1) TN type 1, also referred to as classic TN, where the patient experiences greater episodic pain greater than 50\% of the time; 2) TN type 2, where constant pain occurs greater than 50\% of the time. Trigeminal injury can be unintentional: which can be due to facial trauma, surgery to the oral cavity; this type of pain would involve persistent and burning pain to the affected areas; and is termed trigeminal neuropathic pain. The injury can also be intentional, due to interventions such as rhizotomy or gangliolysis; these result in numbness and loss of sensation of the face and are termed trigeminal deafferentation pain. Symptomatic TN, also termed MS-TN, is due to trigeminal nerve injury relating to multiple sclerosis related central white matter inflammation. Postherpetic TN is pain resulting from the infection of the herpes zoster virus, and can involve burning pain and deafferentation. In this classification, the definition of atypical facial pain is reserved for somatoform pain disorder that can be of psychosomatic origin, although some literature can also include constant pain in the face into this classification. This is in contrast with the common clinical definition, where atypical facial pain is used for the definition of trigeminal pain whose symptomatology deviates from classic TN.

@@

The afferent branches converge and form the trigeminal root ganglion (TRG) (also known as Gasserian, or semilunar ganglion), which then leads to the trunk or cisternal component of CN V. The TRG contains all the trigeminal sensory cell bodies. It situates in the Meckel's cave, in the middle cranial fossa. It is anatomically organized with respect to the trigeminal branches, with V1 situated in medial anterior, V2 in caudal lateral, and V3 in the middle.

@@

 \begin{figure}[ht]
 \includegraphics[width=\textwidth]{trig-pathway-wilson.jpg}
 \centering
 \caption{Overview of the trigeminal afferent pathways \protect\cite{wilson2010cranial}}
 \label{fig:trig-pathway}
 \end{figure}

@@

Janetta \cite{Jannetta1967} proposed that NVC leads to focal demyelination of the trigeminal sensory root axons, and causes ephaptic nerve conductances that cause neuralgia. Ultrastructural examinations \cite{Love1998} of tissue samples obtained from CN V rhizotomy specimens showed oligodendrocyte demyelination at the site of root transition zone. Adjacent to the site of compression-related focal demyelination, thinly myelinated sheaths that encapsulate a number of demyelinated axons could be observed, suggesting aberrant remyelination. It is therefore plausible that NCV disrupts the central myelin. The region of disruption may include the peripheral-central myelin transition zone, and extending into central myelin regions.The TRG is also postulated to be responsible for neuralgia \cite{Rappaport1994}. It is believed that ganglionic ectopic autonomic firing can be triggered by mechanical disturbances as a result of nerve compression. Additionally, ganglionic cross aftercharge may occur as a result of repeated stimulations in the nerve injury site, which ignites surrounding neurons. Histological findings by Devor et al. support the ignition hypothesis \cite{Devor2002a}, where demyelinated axons are shown to be in close apposition without intervening glia.

@@

Intervention can also be non-invasive, in the form of Gamma Knife radiosurgery (GKRS) \cite{Hodaie2012g}. The Perfexion GK system focuses gamma radiation from 192 cobalt-60 ($^{60}$Co) sources onto specific targets along the trigeminal nerve. The target is either on the cisternal segment of CN V or just outside of the REZ \cite{Daugherty2015}. Proximal targets correlate with longer pain relief, but also higher chance of facial numbness\cite{Xu2014}. The procedure requires a stereotactic frame, and a dose of 70-90 Gy delivered to the nerve. The effect is delayed and pain relief can be experienced in 4-6 weeks post-radiosurgery. GKRS treatments result in 75\% initial pain relief \cite{Tuleasca2015,Kondziolka2010}, about 50-80\% remain pain-free at 1 year, 45-65\% 2 years and 30-40\% 3 years after surgery \cite{Kondziolka2010,Sheehan2005}. More favourable treatment outcomes were related to no other accompanying symptoms, no prior surgery and shorter term pain history of fewer than three years.

@@

Application of tractography permitted visual reconstructions of human cranial nerves in vivo \cite{Hodaie2010}. Cisternal CN V could be reliably delineated with Gaussian tensors with just 25 MR gradient directions. This permitted high fidelity visualization and targeted characterization of CN V diffusivity in TN, demonstrating significant reduction in fractional anisotropy (FA) post-surgery at the site of the GK target \cite{Hodaie2009a}. Study of CN V diffusivity in conjunction with deep white matter again revealed reduced FA on the affected CN V, with also bilateral increased in radial, axial, and mean diffusivity. Disruptions were also found in the cognitive-affective, attention, and motor regions of the cortex, consistent with cortical patterns in chronic pain \cite{Desouza2013}. Lowered FA was reported in multiple TN studies \cite{Herweh2007, Lutz2011}. A Study in patients with temporomandibular disorder (TMD) showed a bilateral reduction in CN V FA compared to controls, with the FA of the affected side significantly inverse-correlates with pain duration \cite{Moayedi2012}. This TMD study also revealed similar cognitive and motor disruption, and internal capsule FA significantly inverse-correlated with pain intensity.

>>>>

Modern pain literature continues to update the exact diagnostic definition of TN. The Burchiel classification scheme divides facial pain into 5 major subtypes \cite{Burchiel2003,Eller2005}: idiopathic (without apparent cause), trigeminal injury, symptomatic, postherpetic, and atypical facial pain. Idiopathic TN has 2 sub-types: 1) TN type 1, also referred to as classic TN, where the patient experiences greater episodic pain greater than 50\% of the time; 2) TN type 2, where constant pain occurs greater than 50\% of the time. Trigeminal injury can be unintentional: which can be due to facial trauma, surgery to the oral cavity; this type of pain would involve persistent and burning pain to the affected areas; and is termed trigeminal neuropathic pain. The injury can also be intentional, due to interventions such as rhizotomy or gangliolysis; these result in numbness and loss of sensation of the face and are termed trigeminal deafferentation pain. Symptomatic TN, includes MS-TN, an important form of TN due to trigeminal injury relating to multiple sclerosis related central white matter inflammation and plaques. Postherpetic TN is pain resulting from the infection of the herpes zoster virus, and can involve burning pain and deafferentation. In this classification, the definition of atypical facial pain is reserved for somatoform pain disorder that can be of psychosomatic origin, although some literature can also include constant pain in the face into this classification. This is in contrast with the common clinical definition, where atypical facial pain is used for the definition of trigeminal pain whose symptomatology deviates from classic TN.

@@

The afferent branches converge and form the trigeminal ganglion (TRG) (also known as Gasserian, or semilunar ganglion), which then leads to the trunk or cisternal component of CN V. The TRG contains all the trigeminal sensory cell bodies. It situates in the Meckel's cave, in the middle cranial fossa. It is anatomically organized with respect to the trigeminal branches, with V1 situated in medial anterior, V2 in caudal lateral, and V3 in the middle.

@@

 \begin{figure}[ht]
 \includegraphics[width=110mm]{noback_2005_10_dot_6.jpg}
 \centering
 \caption{Overview of the trigeminal afferent pathways \protect\cite{noback2005human}. Reproduced with permission from Copyright Clearance Center (License ID: 4336271012841).}
 \label{fig:trig-pathway}
 \end{figure}

@@

Janetta \cite{Jannetta1967} proposed that NVC leads to focal demyelination of the trigeminal sensory root axons, and causes ephaptic nerve conductances that cause neuralgia. Ultrastructural examinations \cite{Love1998} of tissue samples obtained from CN V rhizotomy specimens showed oligodendrocyte demyelination at the site of root transition zone. Adjacent to the site of compression-related focal demyelination, thinly myelinated sheaths that encapsulate a number of demyelinated axons could be observed, suggesting aberrant remyelination. It is therefore plausible that NCV disrupts the central myelin. The region of disruption may include the peripheral-central myelin transition zone, and extending into central myelin regions.The TRG is also postulated to be responsible for neuralgia \cite{Rappaport1994}. It is believed that ganglionic ectopic firing can be triggered by mechanical disturbances as a result of nerve compression. Additionally, ganglionic cross action potential discharge may occur as a result of repeated stimulations in the nerve injury site, which ignites surrounding neurons. Histological findings by Devor et al. support the ignition hypothesis \cite{Devor2002a}, where demyelinated axons are shown to be in close apposition without intervening glia.

@@

Intervention can also be non-invasive, in the form of Gamma Knife radiosurgery (GKRS) \cite{Hodaie2012g}. The Perfexion GK system focuses gamma radiation from 192 cobalt-60 ($^{60}$Co) sources onto specific targets along the trigeminal nerve. The target is either on the cisternal segment of CN V or just outside of the REZ \cite{Daugherty2015}. Proximal targets correlate with longer pain relief, but also higher chance of facial numbness\cite{Xu2014}. The procedure requires a stereotactic frame, and a dose of 70-90 Gy delivered to the nerve. The effect is delayed and pain relief can be experienced in 4-6 weeks post-radiosurgery. GKRS treatments result in 75\% initial pain relief \cite{Tuleasca2015,Kondziolka2010}, about 50-80\% remain pain-free at 1 year, 45-65\% 2 years and 30-40\% 3 years after surgery \cite{Kondziolka2010,Sheehan2005}. Better treatment outcomes were related to factors such as having no other accompanying symptoms, no prior surgery and shorter term pain history of fewer than three years.

@@

Tractography allows high quality \textit{in-vivo} reconstructions of human cranial nerves \cite{Hodaie2010}. Cisternal CN V could be reliably delineated with Gaussian tensors with just 25 MR gradient directions. This permitted high fidelity visualization and targeted characterization of CN V diffusivity in TN, demonstrating significant reduction in fractional anisotropy (FA) post-surgery at the site of the GK target \cite{Hodaie2009a}. Study of CN V diffusivity in conjunction with deep white matter again revealed reduced FA on the affected CN V, with also bilateral increased in radial, axial, and mean diffusivity. Disruptions were also found in the cognitive-affective, attention, and motor regions of the cortex, consistent with cortical patterns in chronic pain \cite{Desouza2013}. Lowered FA was reported in multiple TN studies \cite{Herweh2007, Lutz2011}. A study in patients with temporomandibular disorder (TMD) showed a bilateral reduction in CN V FA compared to controls, with the FA of the affected side significantly inverse-correlates with pain duration \cite{Moayedi2012}. This TMD study also revealed similar cognitive and motor disruption, and internal capsule FA significantly inverse-correlated with pain intensity.
